\chosentitle

### 2024-12-20 paper/main.tex

\begin{figure*}
    \includegraphics[width=\textwidth, height=\textwidth]{../code/harps_examples.pdf}
    \caption{Continuum-rectified \harps\ spectra of example spectral types from B to M. The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission. \todo{Only BAF-type stars used blaze-corrected spectra. Need to download blazes for others and re-run.} \ajw{consider a lower line weight} \ajw{I feel like i'm being thick but I don't get whey these ever go above 1. Also, are you sure that the clam continuum is where the theoretical continuum is the the M1 star?}}\label{fig:harps-examples}}
\end{figure*}

@@

There is good agreement between the model fits and data for example stars of different spectral types. In detail there are mis-matches in line absorption strength, or mis-matches in the form of missing lines (in either the model or the data). We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both. In any case, the model appears to be sufficiently accurate to describe the continuum without substantial bias, so it is not a major concern for our purposes.\\

@@

We have shown that a constrained linear model is a good choice for modeling stellar spectra, which enables precise estimates of the continuum for a range of stellar types. Representing stellar absorption using NMF instead of a some other linear representation is particularly advantageous. Non-negativity ensures that stellar basis vectors are strictly additive, and the negative log-transformation we use restricts both the rectified stellar flux and telluric transmission to be between 0 and 1. Any data values outside this range can \emph{only} be captured by the continuum basis vectors. In some sense the predictive capacity of the combination of NMF basis vectors is nearly orthogonal to a combination of the continuum basis vectors.\\

@@

Chosing not to interpret the basis vectors or weights has specific advantages. For example, if we were to build some mapping from stellar parameters to basis weights, then one could construct a forward model for the data where the model parameters are the continuum coefficients and the stellar parameters (e.g., $T_\mathrm{eff}$, $\log{g}$), instead of continuum coefficients and basis weights. That model would be far less flexible than the model we present here. For example, our model is capable of providing excellent fits to the data, even if the underlying theoretical spectra are inaccurate. A simple example is hydrogen lines. \citep{Wheeler} showed that the hydrogen lines in the infrared are incorrectly computed for every spectral synthesis code except for \texttt{Korg}. When we use existing grids (e.g., like those used in \texttt{FERRE}, as part of \texttt{ASPCAP}), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would find a biased estimate of effective temperature, or a worse fit to the data, or both. Instead, by not restricting ourselves to a mapping between stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.\\

@@

The speed of our model makes it particularly advantageous for evaluating on large data sets. If we assume that the factorization of stellar spectra introduces a small error relative to the usual mis-match between models and data, then a constrained linear model represents a promising avenue for identifying regions where models and observations disagree the most. The simplest path for doing this might be to look at the total squared residuals as a function of pixel across many stars, which would immediately indicate the areas where the model does not have capacity to predict the data. Next steps might be to improve modelling for that region (e.g., through better line lists) or might be to iteratively update the basis vectors in a data-driven way so that they better match the data. The model is trained from synthetic spectra, so it is already a close approximation to what stellar spectra look like, and doing data-driven updates would be an excellent way to handle severe data-model mis-matches (e.g., large molecular features which are present in data but we lack line lists to accuratley model). There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we can restrict ourselves to only updating a single basis vector.\\

>>>>

\begin{figure*}
    \includegraphics[width=\textwidth, height=\textwidth]{../code/harps_examples.pdf}
    \caption{Continuum-rectified \harps\ spectra of example spectral types from B to M. The model used here includes 16 basis vectors for stellar absorption, and 4 basis vectors for telluric transmission. \todo{Only BAF-type stars used blaze-corrected spectra. Need to download blazes for others and re-run.} \ajw{consider a lower line weight} \ajw{I feel like i'm being thick but I don't get why these ever go above 1. Also, are you sure that the clam continuum is where the theoretical continuum is for the M1 star?}}\label{fig:harps-examples}}
\end{figure*}

@@

There is good agreement between the model fits and data for example stars of different spectral types. In detail there are mis-matches in line absorption strength, or mis-matches in the form of missing lines (in either the model or the data). We discuss model mismatches in Section~\ref{sec:discussion}, but we can surmise that model mis-matches are a result of missing or inaccurate physics in the spectral synthesis code or its inputs (e.g., model atmospheres, line lists, opacities), or due to the poor linear approximation we make for the stellar spectra, or both. In any case, the model appears to be sufficiently accurate to describe the continuum without substantial bias, so it is not a major concern for our purposes. \\

@@

We have shown that a constrained linear model is a good choice for modeling stellar spectra, which enables precise estimates of the continuum for a range of stellar types. Representing stellar absorption using NMF instead of a some other linear representation is particularly advantageous. Non-negativity ensures that stellar basis vectors are strictly additive, and the negative log-transformation we use restricts both the rectified stellar flux and telluric transmission to be between 0 and 1. Any data values outside this range can \emph{only} be captured by the continuum basis vectors. In some sense the predictive capacity of the combination of NMF basis vectors is nearly orthogonal to a combination of the continuum basis vectors.\ajw{Is this a restatement/interpretation of the fact that the stellar line basis and continuum basis are aproximately orthogonal?}\\

@@

Chosing not to interpret the basis vectors or weights has specific advantages. For example, if we were to build some mapping from stellar parameters to basis weights, then one could construct a forward model for the data where the model parameters are the continuum coefficients and the stellar parameters (e.g., $T_\mathrm{eff}$, $\log{g}$), instead of continuum coefficients and basis weights. That model would be far less flexible than the model we present here. For example, our model is capable of providing excellent fits to the data, even if the underlying theoretical spectra are inaccurate. A simple example is hydrogen lines. \citep{Wheeler} showed that the hydrogen lines in the infrared are incorrectly computed by many spectral synthesis codes. When we use existing grids with this problem (e.g., like those used in \texttt{FERRE}, as part of \texttt{ASPCAP}), if we had a model that was mapping from stellar parameters to basis weights, then even if the hydrogen lines are in their own basis vector, we would find a biased estimate of effective temperature, or a worse fit to the data, or both. Instead, by not restricting ourselves to a mapping between stellar parameters to basis weights, we retain flexibility to make excellent fits to the data.\\

@@

The speed of our model makes it particularly advantageous for evaluating on large data sets. If we assume that the factorization of stellar spectra introduces a small error relative to the usual mis-match between models and data, then a constrained linear model represents a promising avenue for identifying regions where models and observations disagree the most. The simplest path for doing this might be to look at the total squared residuals as a function of pixel across many stars, which would immediately indicate the areas where the model does not have capacity to predict the data. Next steps might be to improve modelling for that region (e.g., through better line lists) or might be to iteratively update the basis vectors in a data-driven way so that they better match the data. The model is trained from synthetic spectra, so it is already a close approximation to what stellar spectra look like, and doing data-driven updates would be an excellent way to handle severe data-model mis-matches (e.g., large molecular features which are present in data but we lack line lists to accuratley model). There is a lot of flexibility when updating the basis vectors using multiplicative update rules: we can restrict ourselves to only allowing updates on particular pixels (where there is sharp disagreement), and we can restrict ourselves to only updating a single basis vector. \ajw{a reader might be puzzled by the claim that you can evaluate the goodness of the model spectra given the discussion above about how it's fine if the model spectra are a bit shit.}\\
